Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold Standard Items

### 2025-06-11 manuscript.qmd
add more references

To demonstrate the validity of psychological measurements and, thus, the practical utility of psychological measurement models, @bortolottiRelevanceAdvantagesUsing2013 compared psychological measurement with physical measurement. They constructed a scale with 27 items intended to assess body height. Using a response format with two response categories, participants were asked to respond to items such as "Do I think, I would do well in a basketball team?" [@bortolottiRelevanceAdvantagesUsing2013, p. 2350]. At the end of the scale, participants were asked to report their actual height in centimeters, the self-reported physical gold standard item. The parameters estimated from the latent variable body height measurement were then compared with the physical, real measurements. The results revealed a strong correlation (.86, not corrected for disattenuation) between the estimated person parameters derived from a two-parameter logistic model and the self-reported physical height. Still, the gold standard did not exhibit the highest loading on the scale [@bortolottiRelevanceAdvantagesUsing2013]. Van der Linden (2016a) also employed a height measurement based on items with two response categories (e.g., "I bump my head quite often."; van der Linden, 2016a, p. 27) to illustrate the usefulness of logistic item response models (for an overview of item response models, see van der Linden, 2016b). According to the author, this approach is unusual, as body height appears to be a physical variable that can only be measured using a yardstick. However, differences in height influence behavior, making it quite plausible to view such behavioral indicators as proxies for body height and to derive a psychological measurement from them. First, gold standard items do not have the highest loadings so far on the latent variable. Second, measuring a "physical" latent variable might be a psychological representation of body height rather than a measure of physical body height.

@@

First, we constructed items for the physical characteristics of body height, body weight, and age. Age was defined as chronological age or age in years, corresponding to the time elapsed since a person's birth (Montepare & Lachman, 1989; Schwall, 2012). Height was defined as the height of an upright person from the sole of the foot to the top of the head in centimeters, and body weight as the physical mass of a person in kilograms (Martin, 1929). The items were developed by 124 psychology students (deductive item construction, Burisch, 1984) and 24 persons (prototype approach, see Broughton, 1984) who deviate from the German population mean at least one standard deviation in the relevant characteristics (separately for women and men). The 24 persons were asked to think of prototypical behaviors for each construct to assess the whole spectrum of the latent variable. This process produced a total of 138 items (response categories were not tested). In a second step, we examined how these items were interpreted and whether they were connected to ideas unrelated to the constructs using cognitive interviews. We conducted eight interviews with a length of two to three hours and applied various cognitive procedures, including probing, paraphrasing, concurrent-think-aloud, and retrospective-think-aloud (Prüfer & Rexroth, 1996; Prüfer & Rexroth, 2005). This should be a sufficient number as recommendations suggest five to 30 interviews and indicate that the most serious problems can already be identified with a small number of interviews (Willis, 2005). The interviews were conducted with people from the target group (four women and four men) aged between 21 and 77 (*M* = 44.00, *SD* = 20.80). The level of education ranged from high school diploma to university degree. After the cognitive interviews, misleading items were reformulated or eliminated (Faulbaum et al., 2009), resulting in 61 remaining items. In a third step, these 61 items were tested online on a new sample, which initially consisted of 456 people. However, we excluded 66 subjects because they had not completed the questionnaire, resulting in a sample of 390 participants aged between 18 and 77 years (*M* = 32.00, *SD* = 12.80), including 302 women (77.4 %) and 88 men (22.6 %). The level of education was distributed as follows: 0.0 % no school leaving certificate, 3.3 % secondary school leaving certificate/elementary school or equivalent, 17.2 % secondary school or equivalent, 35.9 % vocational baccalaureate or high school diploma, 40.8 % college degree or university degree, 2.8 % doctorate or habilitation. Out of all respondents, 95.4 % stated German as their native language. Based on these survey responses, we excluded items if a) their main item loading was not on the intended factor, b) their main and secondary loadings were almost equal, or c) they had loading below or equal to .30. Additionally, if (d) items had very similar content, items with lower loadings were excluded. Based on these criteria 12 items were selected for each construct (see @tbl-5). For each of the three questionnaires, two versions were created – one with a 6-point scale with verbal endpoints ("does not apply" to "applies", only endpoints were labeled) and fully numerical anchoring for the dichotomous response format with "applies" or "does not apply".

@@

Both scale lengths exhibited similar loading patterns, but those with two response categories obtained slightly higher loadings compared to the six response categories in most cases. On average, questionnaires with two response categories showed better descriptive fit indices than those with six. Model fits were within the expected range for questionnaire items (Goretzko et al., 2024). In particular, the fit indices for the questionnaires on body height and weight with two response categories showed the best fit.

@@

When looking at self-reported gender, there are some items with moderate standardized regression weights predicting body height [e.g., item 8 ("Pants are often so short for me.")] and body weight [e.g., item 6 ("I need to lose weight.")]. These items may cause problems when applied to men and women (differential item functioning; see Hilbert et al., 2022, for a comparable case). Finally, there are only non-significant low standardized regression weights of self-reported gender predicting age items.

@@

Gold-standard items did not work as theoretically but empirically expected in this study. However, we still believe they are important. Only the gold standard item for body height – for six, not for two response categories – had its expected highest loading on the latent variable. Bortolotti et al. (2013), who previously applied the same procedure with two response categories to investigate a gold-standard item for body height, found a correlation of .86 with body height. This correlation pretty much matches our correlation with the latent variable. For our 2-response category scale, the correlation was slightly higher (.87), but there were also items which had descriptively equal or higher correlations (biserial correlations) with the sum score than the self-reported physical item (Pearson correlation). This can be due since we reported a latent correlation corrected for diattenuation, and our scale had 12 instead of 27 items. Nevertheless, our results depict a conceptual replication of Bortolotti et al. (2013) with a different questionnaire.

@@

We must be aware, that the model fit does not prove how well the items are formulated (see Pargent et al., 2019). The model fit does also not reveal if the items are helpful to measure the latent variable. The model fit simply compares a model implied variance/covariance matrix with an empirical variance/covariance matrix. In this study, the a priori most suitable items did not have always the highest discrimination parameters or loadings. Thus, simply selecting items according to their loadings and leaving the field to the factor analysis to decide which item is good or not is likely not enough. There must be an item content loading fit.

The reliability estimate of the questionnaire informs us only about the precision of measurement and not whether our measurement is valid. This does not necessarily contradict attempts to choose items based on reliability (e.g., see Zijlmans et al. 2019) since items must have high loadings to achieve high reliability estimates of a scale.

Thus, we suggest using gold standard items as described above. It should be noted that it is critical to implement the gold standard item in the final scale since we then create a part-whole relation. This no desirable situation since all test models (classical an IRT models) assume that the probability of a positive response of a person to an item should not depend on this person’s response to any other item (Debelak & Kollar, 2020).

@@

Previous hints suggest that measuring physical attributes through psychological questionnaires is not always connected to a psychological latent variable guiding behavior (Hilbert et al., 2022; van der Linden, 2016a, b). Similarly, in our study, it remained unclear if a psychological latent variable beyond body height and weight was measured and how such a latent variable could be interpreted. Only for age there was a latent variable, which can be interpreted as cognitive or psychological age (Barak & Schiffman, 1981). Furthermore, it seems helpful to include additional items, that are not intended to be measured but may further explain the item responses of interest, such as gender or items from the nomological net to reveal possible dependencies and to decide whether these dependencies are in line with the definition of the latent variables.

@@

As expected, reliabilities were higher for scales with six response categories compared to those with two. The results regarding to the model fit as well as to validity, indicated as correlations between latent variables compared to the correlation between gold standard items, are mixed. Some of the fit indices turned out to be descriptively better for items with two response categories and some for six. The same holds true for the differences between the correlations of the gold standard items and the correlations of the latent variables. Except for reliability, the study revealed no clear picture to choose two or six response categories. Hilbert et al. (2016, 2022) showed that the same items with a different number of response categories measure not the same latent variable. This result highlights that the choice of the number of response categories constructing a questionnaire should be based on an empirical study comparing several options accompanied by an intensive cognitive pretesting and not only based on the literature which focuses on maximal reliability.

@@

The results presented here prompt reflection and reconsideration of item construction. First, it may be helpful to consider content validity when developing items. Items that either represent a kind of gold standard with specific hypotheses about the ranking of the loadings or construct divergent items that (in)validate the items intended to measure one latent variable should be included in the construction process. Revisiting the example of *warmth* as a personality facet, such additional validation items could, for example, measure warmth itself or openness. Openness as scale or marker items that show high correlations with these warmth items should not be included in the scale. These demands are not entirely new, but newly emphasized by our analyses, calling for more stringent item construction approaches. Second, the response scale choice may not be as clear-cut as previously suggested (e.g., Lee and Paek, 2014). We found that items with two response categories were not per se unfavorable, as often suggested in the literature. While our binary scales obtained lower reliability estimates, as expected (see Preston & Coleman, 2000; Revilla et al., 2014), other results regarding validity and model fit are mixed. Since reliability is easier to control for by adding more suitable items, construct validity seems to be the stronger argument when choosing response categories. In sum, we believe that decisions on response scales should be empirically founded, especially in high stakes situations (e.g., clinical diagnosis or personnel selection), comparing alternative numbers of response categories. In this context, it is important to recognize that the number of response categories can influence what latent variables are measured (see Hilbert et al. 2016; 2022). Thus, comparing the construct validity of questionnaires differing in the number of response categories is essential. Nevertheless, it remains necessary to subject response scales to intensive cognitive pretesting to determine a suitable number of categories. On the one hand, cognitive pretesting can reveal when single categories are systematically overlooked by respondents lacking information or cognitive capacity to differentiate categories. On the other hand, cognitive pretests help assess participant burden, which should always be kept in mind. For example, two-point scales, which are commonly criticized, have previously been shown to require less effort from the respondents (Hilbert et al, 2016).

>>>>

To demonstrate the validity of psychological measurements and, thus, the practical utility of psychological measurement models, @bortolottiRelevanceAdvantagesUsing2013 compared psychological measurement with physical measurement. They constructed a scale with 27 items intended to assess body height. Using a response format with two response categories, participants were asked to respond to items such as "Do I think, I would do well in a basketball team?" [@bortolottiRelevanceAdvantagesUsing2013, p. 2350]. At the end of the scale, participants were asked to report their actual height in centimeters, the self-reported physical gold standard item. The parameters estimated from the latent variable body height measurement were then compared with the physical, real measurements. The results revealed a strong correlation (.86, not corrected for disattenuation) between the estimated person parameters derived from a two-parameter logistic model and the self-reported physical height. Still, the gold standard did not exhibit the highest loading on the scale [@bortolottiRelevanceAdvantagesUsing2013]. Van der @vanderlindenHandbookItemResponse2016 also employed a height measurement based on items with two response categories (e.g., "I bump my head quite often."; van der Linden, 2016a, p. 27) to illustrate the usefulness of logistic item response models (for an overview of item response models, see van der Linden, 2016b). According to the author, this approach is unusual, as body height appears to be a physical variable that can only be measured using a yardstick. However, differences in height influence behavior, making it quite plausible to view such behavioral indicators as proxies for body height and to derive a psychological measurement from them. First, gold standard items do not have the highest loadings so far on the latent variable. Second, measuring a "physical" latent variable might be a psychological representation of body height rather than a measure of physical body height.

@@

First, we constructed items for the physical characteristics of body height, body weight, and age. Age was defined as chronological age or age in years, corresponding to the time elapsed since a person's birth [@montepareYouReOnly1989; @schwallDefiningAgeUsing2012]. Height was defined as the height of an upright person from the sole of the foot to the top of the head in centimeters, and body weight as the physical mass of a person in kilograms [@martinAnthropometrieAnleitungSelbstaendigen1929]. The items were developed by 124 psychology students [deductive item construction, @burischApproachesPersonalityInventory1984] and 24 persons [prototype approach, see @broughtonPrototypeStrategyConstruction1984] who deviate from the German population mean at least one standard deviation in the relevant characteristics (separately for women and men). The 24 persons were asked to think of prototypical behaviors for each construct to assess the whole spectrum of the latent variable. This process produced a total of 138 items (response categories were not tested). In a second step, we examined how these items were interpreted and whether they were connected to ideas unrelated to the constructs using cognitive interviews. We conducted eight interviews with a length of two to three hours and applied various cognitive procedures, including probing, paraphrasing, concurrent-think-aloud, and retrospective-think-aloud [@prueferVerfahrenEvaluationSurvey1996; @prueferKognitiveInterviews2005]. This should be a sufficient number as recommendations suggest five to 30 interviews and indicate that the most serious problems can already be identified with a small number of interviews [@willisCognitiveInterviewing2005]. The interviews were conducted with people from the target group (four women and four men) aged between 21 and 77 (*M* = 44.00, *SD* = 20.80). The level of education ranged from high school diploma to university degree. After the cognitive interviews, misleading items were reformulated or eliminated [@faulbaumWasIstGute2009], resulting in 61 remaining items. In a third step, these 61 items were tested online on a new sample, which initially consisted of 456 people. However, we excluded 66 subjects because they had not completed the questionnaire, resulting in a sample of 390 participants aged between 18 and 77 years (*M* = 32.00, *SD* = 12.80), including 302 women (77.4 %) and 88 men (22.6 %). The level of education was distributed as follows: 0.0 % no school leaving certificate, 3.3 % secondary school leaving certificate/elementary school or equivalent, 17.2 % secondary school or equivalent, 35.9 % vocational baccalaureate or high school diploma, 40.8 % college degree or university degree, 2.8 % doctorate or habilitation. Out of all respondents, 95.4 % stated German as their native language. Based on these survey responses, we excluded items if a) their main item loading was not on the intended factor, b) their main and secondary loadings were almost equal, or c) they had loading below or equal to .30. Additionally, if (d) items had very similar content, items with lower loadings were excluded. Based on these criteria 12 items were selected for each construct (see @tbl-5). For each of the three questionnaires, two versions were created – one with a 6-point scale with verbal endpoints ("does not apply" to "applies", only endpoints were labeled) and fully numerical anchoring for the dichotomous response format with "applies" or "does not apply".

@@

Both scale lengths exhibited similar loading patterns, but those with two response categories obtained slightly higher loadings compared to the six response categories in most cases. On average, questionnaires with two response categories showed better descriptive fit indices than those with six. Model fits were within the expected range for questionnaire items [@goretzkoEvaluatingModelFit2024]. In particular, the fit indices for the questionnaires on body height and weight with two response categories showed the best fit.

@@

When looking at self-reported gender, there are some items with moderate standardized regression weights predicting body height [e.g., item 8 ("Pants are often so short for me.")] and body weight [e.g., item 6 ("I need to lose weight.")]. These items may cause problems when applied to men and women [differential item functioning\; see @hilbertWhatMeasureEmpirical2022, for a comparable case]. Finally, there are only non-significant low standardized regression weights of self-reported gender predicting age items.

@@

Gold-standard items did not work as theoretically but empirically expected in this study. However, we still believe they are important. Only the gold standard item for body height – for six, not for two response categories – had its expected highest loading on the latent variable. @bortolottiRelevanceAdvantagesUsing2013, who previously applied the same procedure with two response categories to investigate a gold-standard item for body height, found a correlation of .86 with body height. This correlation pretty much matches our correlation with the latent variable. For our 2-response category scale, the correlation was slightly higher (.87), but there were also items which had descriptively equal or higher correlations (biserial correlations) with the sum score than the self-reported physical item (Pearson correlation). This can be due since we reported a latent correlation corrected for diattenuation, and our scale had 12 instead of 27 items. Nevertheless, our results depict a conceptual replication of @bortolottiRelevanceAdvantagesUsing2013 with a different questionnaire.

@@

We must be aware, that the model fit does not prove how well the items are formulated [see @pargentCanMakeIt2019]. The model fit does also not reveal if the items are helpful to measure the latent variable. The model fit simply compares a model implied variance/covariance matrix with an empirical variance/covariance matrix. In this study, the a priori most suitable items did not have always the highest discrimination parameters or loadings. Thus, simply selecting items according to their loadings and leaving the field to the factor analysis to decide which item is good or not is likely not enough. There must be an item content loading fit.

The reliability estimate of the questionnaire informs us only about the precision of measurement and not whether our measurement is valid. This does not necessarily contradict attempts to choose items based on reliability [e.g., see @zijlmansItemScoreReliabilitySelection2019] since items must have high loadings to achieve high reliability estimates of a scale.

Thus, we suggest using gold standard items as described above. It should be noted that it is critical to implement the gold standard item in the final scale since we then create a part-whole relation. This no desirable situation since all test models (classical and IRT models) assume that the probability of a positive response of a person to an item should not depend on this person’s response to any other item [@debelakTestingLocalIndependence2020].

@@

Previous hints suggest that measuring physical attributes through psychological questionnaires is not always connected to a psychological latent variable guiding behavior (@hilbertWhatMeasureEmpirical2022; van der Linden, 2016a, b). Similarly, in our study, it remained unclear if a psychological latent variable beyond body height and weight was measured and how such a latent variable could be interpreted. Only for age there was a latent variable, which can be interpreted as cognitive or psychological age [@barakCognitiveAgeNonchronological1981]. Furthermore, it seems helpful to include additional items, that are not intended to be measured but may further explain the item responses of interest, such as gender or items from the nomological net to reveal possible dependencies and to decide whether these dependencies are in line with the definition of the latent variables.

@@

As expected, reliabilities were higher for scales with six response categories compared to those with two. The results regarding to the model fit as well as to validity, indicated as correlations between latent variables compared to the correlation between gold standard items, are mixed. Some of the fit indices turned out to be descriptively better for items with two response categories and some for six. The same holds true for the differences between the correlations of the gold standard items and the correlations of the latent variables. Except for reliability, the study revealed no clear picture to choose two or six response categories. Hilbert et al. [-@hilbertInfluenceResponseFormat2016; -@hilbertWhatMeasureEmpirical2022] showed that the same items with a different number of response categories measure not the same latent variable. This result highlights that the choice of the number of response categories constructing a questionnaire should be based on an empirical study comparing several options accompanied by an intensive cognitive pretesting and not only based on the literature which focuses on maximal reliability.

@@

The results presented here prompt reflection and reconsideration of item construction. First, it may be helpful to consider content validity when developing items. Items that either represent a kind of gold standard with specific hypotheses about the ranking of the loadings or construct divergent items that (in)validate the items intended to measure one latent variable should be included in the construction process. Revisiting the example of *warmth* as a personality facet, such additional validation items could, for example, measure warmth itself or openness. Openness as scale or marker items that show high correlations with these warmth items should not be included in the scale. These demands are not entirely new, but newly emphasized by our analyses, calling for more stringent item construction approaches. Second, the response scale choice may not be as clear-cut as previously suggested [e.g., @leeSearchOptimalNumber2014]. We found that items with two response categories were not per se unfavorable, as often suggested in the literature. While our binary scales obtained lower reliability estimates, as expected [see @prestonOptimalNumberResponse2000; @revillaChoosingNumberCategories2014], other results regarding validity and model fit are mixed. Since reliability is easier to control for by adding more suitable items, construct validity seems to be the stronger argument when choosing response categories. In sum, we believe that decisions on response scales should be empirically founded, especially in high stakes situations (e.g., clinical diagnosis or personnel selection), comparing alternative numbers of response categories. In this context, it is important to recognize that the number of response categories can influence what latent variables are measured [see @hilbertInfluenceResponseFormat2016; @hilbertWhatMeasureEmpirical2022]. Thus, comparing the construct validity of questionnaires differing in the number of response categories is essential. Nevertheless, it remains necessary to subject response scales to intensive cognitive pretesting to determine a suitable number of categories. On the one hand, cognitive pretesting can reveal when single categories are systematically overlooked by respondents lacking information or cognitive capacity to differentiate categories. On the other hand, cognitive pretests help assess participant burden, which should always be kept in mind. For example, two-point scales, which are commonly criticized, have previously been shown to require less effort from the respondents [@hilbertInfluenceResponseFormat2016].
